---
title:  "EDRM 718 Sp21: Exercise Set C"
author:  "Geophrey O. Odero"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_notebook
---


```{r include=FALSE}
# these libraries are good when it comes to plots and data manipulation

library(here)
library(plyr)
library(gtools)
library(ggplot2)
```

*****
**Directions:** _:Complete this assignment in an R notebook. Make sure to number each item, but it is optional for you to copy the prompt for the item. Make sure that your notebook is neat and organized and that all your code chunks run without error, even after clearing the environment. Include both the exercise set letter and your name in your notebook filename. When you are finished, push your repository to GitHub. (Four points per item unless stated otherwise.)_

*****

Question #1). **Create a list, my.courses, that contains three vectors: one with the numbers (e.g. EDRM 718) of the courses you are taking this semester, one with the names of the instructors of these courses (in order), and one with the credits you’ll earn in the course (again, in order).**

*****

```{r}
# Load data--------------

hsb2<-here("Data","hsb2.csv")
```

QUestion #2). **Use the HSB2 data. Store test score matrices (include all five test scores) by SES in a list. Your matrix names should be indicative of SES level so that when you look at the list you can tell which matrix corresponds to each SES level. Once your list is set up, use it to plot the relationship of math score (response variable, so placed in the vertical direction) to reading score (explanatory variable) for each SES level. Make sure to label your axes. Also provide a title that clearly identifies each plot. Your plot should also include text that provides the correlation among the variables. (Note: Your script should make this automatic. You should not have to change the script when the correlation changes.) (25 points)**


*****

Question #3). **Use the sample function to randomly select three classmates. Consider these classmates and  yourself as a learning group of four students. Create a data frame for your learning group that includes the following variables: first name, last name, email address, program of study, number of years in current program of study. (You can email your selected classmates for this information. You will find email addresses on a previous Research Café post.) After the data frame is created, add two more variables: one that is the name again, but in “first name, last name” format (this variable should be created from your original data frame and should be placed right after the first name and last name variables) and a logical (also created from existing data and not “by hand”) to indicate whether the group member has been in their program of study more than three years. (10 points) **

*****

Question #4). **Write a script that (a) loads the Mount Juniper Elementary data into a data frame, (b) checks the score column to determine if there are missing data, and (c) prints either there are no missing data or there are missing data, depending on which of these is true.**

*****

Question #5). **Write a script to conduct an analysis of variance to determine if you can infer different score means for the three different grade levels at Mount Juniper Elementary. (Caution: Look at the degrees of freedom for grade level to make certain you are doing this correctly.) (10 points)**

*****

Question #6). **Add lines to your script to store your ANOVA results in an object and then examine your ANOVA results and correctly print “Statistically significant” or “Not statistically significant.” Use $\alpha = 0.05$. (Note: Avoid showing a warning message. If you need help, head to the Research Café.) (10 points)**

*****

Question #7). **Add lines to your script to add a “Results” attribute to your object that contains the words “Statistically significant” or “Not statistically significant.” Use α = 0.05. (Again, avoid a warning message.)**

*****

Question #8). **Consider a population of normally distributed scores with a mean of 100 and a standard deviation of 15. Write a script to draw 1,000 random samples, each of size 40, from this population. Calculate and store each mean in a vector of means. Calculate the mean of the sample means. Determine the difference between the true mean of 100 and the mean of the sample means. What statistical characteristic of the mean does this illustrate? (10 points) **

****

Question #9). **Use the HSB2 data to illustrate how each member of the apply family works. Do this by selecting the correct family member to accomplish each of the following tasks for the five test scores (not the entire data set). Make sure to use each family member one time, and only one time. (10 points)**

a. Put the standard deviation (SD) of each test as an element in a list of the five SDs.
b. Create a vector of SDs for the five tests.
c. Create a vector of the standard deviation of the five test scores for each student in the sample.
d. Obtain SDs for the five tests by school type (i.e., public or private)

****


Question #10). **Construct a vector of any nine different numbers between 0 and 100. Next, add two missing indicators to your vector, somewhere mixed in among the numbers. Now add infinity to your vector, again anywhere within the vector. You should now have a vector of length 12. Use your vector to demonstrate, with explanation, all possible arguments for the range function. If the range function can be used on more types of data than numeric, also construct additional vectors to demonstrate all possible uses of the range function. (13 points)**

























